LUCID: Learning-Enabled Uncertainty-Aware Certification of Stochastic Dynamical Systems

### 2025-08-01 docs/paper/appendix.tex
feat: move to appendix

\pylucid provides a \gls{gui} to guide the user in the creation of the scenario configuration and presenting the results in an intuitive way.
Running \lstinline|pylucid-gui| will open a browser tab containing the interface shown in Figures~\ref{fig:gui-left}--\ref{fig:gui-right}, while a local server will listen for requests coming from the \gls{gui}, computing and returning the results.
\begin{figure}
    \centering
    \includegraphics[trim={0 12.4cm 15cm 0},clip,width=.7\linewidth]{figures/frontend.png}
    \caption{\lucid's \gls{gui} can be accessed via any web browser. 
    The left half of the screen allows the user to provide the transition samples and the safety specification to verify.
    }
    \label{fig:gui-left}
\end{figure}
\begin{figure}
    \centering
    \includegraphics[trim={0 12.4cm 15cm 0},clip,width=.7\linewidth]{figures/frontend.png}
    \caption{\lucid's \gls{gui} can be accessed via any web browser. 
    The right half of the screen shows the results produced by \lucid, including a plot of the computed barrier function, 
    the numerical safety guarantee and , and detailed logs of the computations.
    }
    \label{fig:gui-right}
\end{figure}

@@

We consider the following system:
\begin{equation*}
      \begin{bmatrix}
            {x}_{1, t+1} \\
            {x}_{2, t+1}
      \end{bmatrix}
      = \begin{bmatrix}
            {x}_{2, t} - 1 + e^{-x_{1, t}} \\
            -\sin^2(x_{1, t})
      \end{bmatrix} + w_t,
\end{equation*}
where $w_t\sim\mathcal{N}(\cdotx\vert 0,0.01I_2)$.
Given
\begin{align*}
       & \X = [ -2, 2 ] \times [ -2 , 2 ]                                  \\
       & \X_0 = \{ [x_1, x_2] : (x_1 + 0.5)^2 + (x_2 - 0.5)^2 \leq 0.4 \}  \\
       & \X_U = \{ [x_1, x_2] : (x_1 - 0.7)^2 + (x_2 + 0.7)^2 \leq 0.3 \},
\end{align*}
we want to ensure that the system, starting in $\X_0$, does not enter the unsafe regions $\X_U$ within $T=5$ time steps.
The complete configuration for the \barrII benchmark is shown in Listing~\ref{lst:barrier2}.
\lstinputlisting[language=yaml,caption={Configuration for \barrII},captionpos=b,label={lst:barrier2}]{code/barrier2.yaml}

@@

We consider the following system:
\begin{equation*}
      \begin{bmatrix}
            {x}_{1, t+1} \\
            {x}_{2, t+1}
      \end{bmatrix}
      = \begin{bmatrix}
            {x}_{2, t} \\
            \frac{1}{3} {x}^3_{1, t} - {x}_{1,t} - {x}_{2,t}
      \end{bmatrix} + w_t,
\end{equation*}
where $w_t\sim\mathcal{N}(\cdotx\vert 0,0.01I_2)$.
Given
\begin{align*}
       & \X = [ -3, 2.5 ] \times [ -2 , 1 ]                                                 \\
       & \X_0 = [ 1 , 2 ] \times [ -0.7 , 0.3 ] \cup [ -1.8 , -1.4 ] \times [ -0.1 , 0.1 ]  \\
       & \qquad \cup [-1.4, -1.2] \times [-0.5 , 0.1]                                       \\
       & \X_U = [ 0.4 , 0.6 ] \times [ 0.2 , 0.6 ] \cup [ 0.6 , 0.7 ] \times [ 0.2 , 0.4 ],
\end{align*}
we want to ensure that the system, starting in $\X_0$, does not enter the unsafe regions $\X_U$ within $T=5$ time steps.
The complete configuration for the \barrIII benchmark is shown in Listing~\ref{lst:barrier3}.
\lstinputlisting[language=yaml,caption={Configuration for \barrIII},captionpos=b,label={lst:barrier3}]{code/barrier3.yaml}

@@

We consider a scenario where an autonomous vehicle (AV) controlled by a \gls{nn} is overtaking another vehicle.
The dynamics of the ego vehicle are given by Dubin's car model \EC{does it need a cit or is it just textbook} by adding the noise vector $w = \begin{bmatrix}w_t^1 & w_t^2 & w_t^3\end{bmatrix}$ where each component is drawn from a zero-mean Gaussian with standard deviation $0.01$, $0.01$, and $0.001$, respectively.
The steering wheel angle is supplied by the \gls{nn} controller and we travel at a fixed velocity $v=1$.
Given
\begin{align*}
       & \X = [ 1, 90 ] \times [ -7, 19 ] \times [ -\pi, \pi ]         \\
       & \X_0 = [ 1, 3 ] \times [ -1, 1 ] \times [ -0.5, 0.5 ]         \\
       & \X_U = [ 1, 90 ] \times [ -7, -6 ] \times [ -\pi, \pi ]       \\
       & \qquad \cup [ 1, 90 ] \times [ 18, 19 ] \times [ -\pi, \pi ]  \\
       & \qquad \cup [ 40, 45 ] \times [ -6, 6 ] \times [ -\pi, \pi ],
\end{align*}
we want to ensure that the system, starting in $\X_0$, does not enter the unsafe regions $\X_U$ within $T=5$ time steps.

@@

>>>>

\pylucid provides a \gls{gui} to guide the user in the creation of the scenario configuration and presenting the results in an intuitive way.
Running \lstinline|pylucid-gui| will open a browser tab containing the interface shown in Figures~\ref{fig:gui-left}--\ref{fig:gui-right}, while a local server will listen for requests coming from the \gls{gui}, computing and returning the results.
Following the same numeration as the arrows in the figures, we provide a list of the main components of the \gls{gui}:
\begin{enumerate}
      \item Only a subsection of the configuration options is shown at any given moment. More options can be accessed by clicking on the \textit{Advanced} button.
      \item It is easy to import or export the configuration using the Json format. The experiments discussed in this paper are also available in the \textit{Import} menu.
      \item Clicking the \textit{Preview} button produces a visualization of the system function described by the transition samples.
      \item The dimension of the system can be specified in the \textit{Model} tab.
      \item You can either provide the transition samples yourself in the \textit{Data} tab, or specify a transition function in the \textit{Model} tab.
      \item The $x^1,\ldots,x^N$ samples, stored in \texttt{csv} format.
      \item The $x^1_+,\ldots,x^N_+$ samples, stored in \texttt{csv} format.
      \item The bounding box of the state space.
      \item The initial set $\X_0$.
      \item The unsafe set $\X_U$.
      \item Clicking the \textit{Submit} button will start the computation of the barrier function with the current configuration.
      \item Plot of the transition function or the computed barrier function.
      \item Numerical results of the computation.
      \item Logs of the computation.
\end{enumerate}
\begin{figure}
      \centering
      \includegraphics[trim={0 0 17cm 0},clip,width=.7\linewidth]{figures/frontend-left.png}
      \caption{The left half of the page allows the user to provide the transition samples and the safety specification to verify.
      }
      \label{fig:gui-left}
\end{figure}
\begin{figure}
      \centering
      \includegraphics[trim={13.5cm 12.4cm 0 0},clip,width=.7\linewidth]{figures/frontend-right.png}
      \caption{The right half of the page shows the results produced by \lucid, including a plot of the computed barrier function and detailed logs.}
      \label{fig:gui-right}
\end{figure}

@@

We consider the following system:
\begin{equation*}
      \begin{bmatrix}
            {x}_{1, t+1} \\
            {x}_{2, t+1}
      \end{bmatrix}
      = \begin{bmatrix}
            {x}_{2, t} - 1 + e^{-x_{1, t}} \\
            -\sin^2(x_{1, t})
      \end{bmatrix} + w_t,
\end{equation*}
where $w_t\sim\mathcal{N}(\cdotx\vert 0,0.01I_2)$.
Given
\begin{align*}
       & \X = [ -2, 2 ] \times [ -2 , 2 ]                                  \\
       & \X_0 = \{ [x_1, x_2] : (x_1 + 0.5)^2 + (x_2 - 0.5)^2 \leq 0.4 \}  \\
       & \X_U = \{ [x_1, x_2] : (x_1 - 0.7)^2 + (x_2 + 0.7)^2 \leq 0.3 \},
\end{align*}
we want to ensure that the system, starting in $\X_0$, does not enter the unsafe regions $\X_U$ within $T=5$ time steps.
The kernel \hp set to be $\sigma_f=1$ and $\sigma_l=[1.0, 5.39334]$.
Using a spectral basis of $M=15^2-1$ frequencies and a lattice density of $\hat{N} = 350^2$,
we were able to generate a \gls{cbc} with $\eta \approx 5.032$, $\gamma = 100$, and $c \approx 3.926$ in $14.7$ minutes, achieving a lower bound on the safety probability of $P_{\text{safe}}(\S^\pi) \geq 75.34\%$.
The result is shown in Figure~\ref{fig:CSBarr2}.
The complete configuration for the \barrII benchmark is shown in Listing~\ref{lst:barrier2}.
\lstinputlisting[language=yaml,caption={Configuration for \barrII},captionpos=b,label={lst:barrier2}]{code/barrier2.yaml}

@@

We consider the following system:
\begin{equation*}
      \begin{bmatrix}
            {x}_{1, t+1} \\
            {x}_{2, t+1}
      \end{bmatrix}
      = \begin{bmatrix}
            {x}_{2, t} \\
            \frac{1}{3} {x}^3_{1, t} - {x}_{1,t} - {x}_{2,t}
      \end{bmatrix} + w_t,
\end{equation*}
where $w_t\sim\mathcal{N}(\cdotx\vert 0,0.01I_2)$.
Given
\begin{align*}
       & \X = [ -3, 2.5 ] \times [ -2 , 1 ]                                                 \\
       & \X_0 = [ 1 , 2 ] \times [ -0.7 , 0.3 ] \cup [ -1.8 , -1.4 ] \times [ -0.1 , 0.1 ]  \\
       & \qquad \cup [-1.4, -1.2] \times [-0.5 , 0.1]                                       \\
       & \X_U = [ 0.4 , 0.6 ] \times [ 0.2 , 0.6 ] \cup [ 0.6 , 0.7 ] \times [ 0.2 , 0.4 ],
\end{align*}
we want to ensure that the system, starting in $\X_0$, does not enter the unsafe regions $\X_U$ within $T=5$ time steps.
The kernel \hp were set to $\sigma_f=1$ and $\sigma_l=[1.0, 1.0, 1.0]$.
Hence, we need a rather high number of frequencies to achieve the required expressivity in the barrier.
Using a spectral basis of $M=15^3-1$ frequencies and a lattice density of $\hat{N} = 30^2$,
we were able to generate a \gls{cbc} with $\eta \approx 0.1$, $\gamma = 100$, and $c \approx 0.1$ in $XX:XX$ minutes, achieving a lower bound on the safety probability of $P_{\text{safe}}(\S^\pi) \geq XX\%$.
The result is shown in Figure~\ref{fig:CSBarr3}.
The complete configuration for the \barrIII benchmark is shown in Listing~\ref{lst:barrier3}.
\lstinputlisting[language=yaml,caption={Configuration for \barrIII},captionpos=b,label={lst:barrier3}]{code/barrier3.yaml}

@@

We consider a scenario where an autonomous vehicle (AV) controlled by a \gls{nn} is overtaking another vehicle.
The dynamics of the ego vehicle are given by Dubin's car model by adding the noise vector $w = \begin{bmatrix}w_t^1 & w_t^2 & w_t^3\end{bmatrix}$ where each component is drawn from a zero-mean Gaussian with standard deviation $0.01$, $0.01$, and $0.001$, respectively.
The steering wheel angle is supplied by the \gls{nn} controller and we travel at a fixed velocity.
Given
\begin{align*}
       & \X = [ 1, 90 ] \times [ -7, 19 ] \times [ -\pi, \pi ]         \\
       & \X_0 = [ 1, 3 ] \times [ -1, 1 ] \times [ -0.5, 0.5 ]         \\
       & \X_U = [ 1, 90 ] \times [ -7, -6 ] \times [ -\pi, \pi ]       \\
       & \qquad \cup [ 1, 90 ] \times [ 18, 19 ] \times [ -\pi, \pi ]  \\
       & \qquad \cup [ 40, 45 ] \times [ -6, 6 ] \times [ -\pi, \pi ],
\end{align*}
we want to ensure that the system, starting in $\X_0$, does not enter the unsafe regions $\X_U$ within $T=5$ time steps.

@@

% Linear
% Success: 93.74%, c 0.0032813476604243406, eta 0.013421457997340467, lambda 1e-05, num_frequencies 5, num_oversample 704, oversample_factor 2.0, sigma_l 0.034, sigma_f 18.0, T 15
% \begin{tabular}{rlrrrrrrrrl}
%   \toprule
%   sigma_f   & sigma_l & lambda_  & num_frequencies & num_oversample & T  & gamma    & eta      & c        & percentage & time \\
%   \midrule
%   18.000000 & 0.034   & 0.000010 & 5               & 704            & 15 & 1.000000 & 0.013421 & 0.003281 & 93.735833  & 0:01 \\
%   \bottomrule
% \end{tabular}

% Barrier 2
% \begin{tabular}{rlrrrrrrrrl}
%     \toprule
%     sigma_f  & sigma_l           & lambda_  & num_frequencies & num_oversample & T & gamma      & eta      & c        & percentage & time  \\
%     \midrule
%     1.000000 & [2.50304 3.77779] & 0.000001 & 15              & 400            & 5 & 2.000000   & 0.155993 & 0.141315 & 56.871502  & 10:01 \\
%     1.000000 & [2.50304 3.77779] & 0.000001 & 9               & 350            & 5 & 100.000000 & 8.116091 & 7.537732 & 54.195247  & 2:02  \\
%     1.000000 & [2.50304 3.77779] & 0.000001 & 11              & 300            & 5 & 2.000000   & 0.183647 & 0.150238 & 53.258213  & 2:25  \\
%     1.000000 & [2.50304 3.77779] & 0.000001 & 13              & 300            & 5 & 2.000000   & 0.182542 & 0.150821 & 53.167620  & 3:53  \\
%     1.000000 & [2.50304 3.77779] & 0.000001 & 9               & 300            & 5 & 2.000000   & 0.176700 & 0.153188 & 52.868065  & 1:28  \\
%     1.000000 & [2.50304 3.77779] & 0.000001 & 9               & 300            & 5 & 100.000000 & 8.835006 & 7.659386 & 52.868065  & 1:33  \\
%     1.000000 & [2.50304 3.77779] & 0.000001 & 20              & 300            & 5 & 2.000000   & 0.223455 & 0.147586 & 51.930650  & 16:48 \\
%     1.000000 & [2.50304 3.77779] & 0.000001 & 9               & 250            & 5 & 100.000000 & 9.875340 & 7.785029 & 51.199515  & 1:10  \\
%     \bottomrule
% \end{tabular}

% Barrier3
% Success: 42.25%, c 0.11928419647787443, eta 0.5584972599863635, lambda 1e-08, num_frequencies 15, num_oversample 800, oversample_factor 1.0, sigma_l [2.99266 4.62946], sigma_f 1.0, T 5
% \begin{tabular}{rlrrrrrrrrl}
%     \toprule
%     sigma_f  & sigma_l           & lambda_  & num_frequencies & num_oversample & T & gamma    & eta      & c        & percentage & time  \\
%     \midrule
%     1.000000 & [2.99266 4.62946] & 0.000000 & 15              & 800            & 5 & 2.000000 & 0.558497 & 0.119284 & 42.254088  & 64:20 \\
%     1.000000 & [2.99266 4.62946] & 0.000000 & 15              & 600            & 5 & 2.000000 & 0.598432 & 0.122033 & 39.570103  & 25:12 \\
%     1.000000 & [2.99266 4.62946] & 0.000000 & 15              & 450            & 5 & 2.000000 & 0.638667 & 0.127644 & 36.155716  & 12:07 \\
%     1.000000 & [2.99266 4.62946] & 0.000000 & 14              & 450            & 5 & 2.000000 & 0.638151 & 0.128054 & 36.078977  & 9:58  \\
%     1.000000 & [2.99266 4.62946] & 0.000000 & 14              & 400            & 5 & 2.000000 & 0.651139 & 0.130655 & 34.779342  & 7:50  \\
%     1.000000 & [2.99266 4.62946] & 0.000000 & 14              & 370            & 5 & 2.000000 & 0.662921 & 0.133060 & 33.589019  & 6:49  \\
%     1.000000 & [2.99266 4.62946] & 0.000000 & 16              & 370            & 5 & 2.000000 & 0.646389 & 0.136513 & 33.552259  & 11:35 \\
%     1.000000 & [2.99266 4.62946] & 0.000000 & 12              & 370            & 5 & 2.000000 & 0.710767 & 0.127533 & 32.578438  & 4:38  \\
%     1.000000 & [2.99266 4.62946] & 0.000000 & 11              & 350            & 5 & 2.000000 & 0.755459 & 0.128623 & 30.071348  & 3:14  \\
%     1.000000 & [2.99266 4.62946] & 0.000000 & 10              & 350            & 5 & 2.000000 & 0.812608 & 0.125781 & 27.924451  & 2:34  \\
%     1.000000 & [2.99266 4.62946] & 0.000000 & 9               & 350            & 5 & 2.000000 & 0.836442 & 0.128851 & 25.965203  & 2:02  \\
%     \bottomrule
% \end{tabular}

% Overtaking
% Success: 30.32%, c 0.05031756638044985, eta 0.09681458793205037, lambda 1e-05, num_frequencies 5, num_oversample -1, oversample_factor 9.0, sigma_l [10.  7.  5.], sigma_f 7.0, T 5
% \begin{tabular}{rlrrrrrrrrl}
%     \toprule
%     sigma_f  & sigma_l       & lambda_  & num_frequencies & num_oversample & T & gamma    & eta      & c        & percentage & time  \\
%     \midrule
%     7.000000 & [10.  7.  5.] & 0.000010 & 5               & 99             & 5 & 0.500000 & 0.096815 & 0.050318 & 30.319516  & 46:33 \\
%     7.000000 & [10.  7.  5.] & 0.000010 & 5               & 100            & 5 & 0.500000 & 0.100419 & 0.050519 & 29.396944  & 49:21 \\
%     7.000000 & [10.  7.  5.] & 0.000010 & 5               & 88             & 5 & 0.500000 & 0.093582 & 0.052546 & 28.737764  & 25:52 \\
%     7.000000 & [10.  7.  5.] & 0.000010 & 5               & 77             & 5 & 0.500000 & 0.103997 & 0.054252 & 24.948829  & 13:49 \\
%     7.000000 & [10.  7.  5.] & 0.000010 & 5               & 66             & 5 & 0.500000 & 0.124539 & 0.058476 & 16.616154  & 8:16  \\
%     7.000000 & [10.  7.  5.] & 0.000010 & 4               & 90             & 5 & 0.500000 & 0.101765 & 0.067857 & 11.790251  & 11:21 \\
%     7.000000 & [10.  7.  5.] & 0.000010 & 4               & 81             & 5 & 0.500000 & 0.109691 & 0.068535 & 9.527199   & 8:08  \\
%     7.000000 & [10.  7.  5.] & 0.000010 & 5               & 55             & 5 & 0.500000 & 0.135738 & 0.063642 & 9.210252   & 4:15  \\
%     7.000000 & [10.  7.  5.] & 0.000010 & 4               & 72             & 5 & 0.500000 & 0.109150 & 0.069704 & 8.465822   & 5:10  \\
%     7.000000 & [10.  7.  5.] & 0.000010 & 5               & 44             & 5 & 0.500000 & 0.160935 & 0.062971 & 4.842390   & 2:06  \\
%     7.000000 & [10.  7.  5.] & 0.000010 & 4               & 63             & 5 & 0.500000 & 0.120951 & 0.071611 & 4.198906   & 3:25  \\
%     \bottomrule
% \end{tabular}

### 2025-08-01 docs/paper/figures/architecture.tex
feat: move to appendix

        % Lucid blocks
        \node[block, below=1cm of input] (estimator) {\Circled{2} Estimator};
        \node[block, right=of estimator] (tuner) {\Circled{3} Tuner};
        \node[block, left=of estimator] (sets) {\Circled{1} Safety Specification};
        \node[block, below=of estimator, text width=2.4cm] (feature) {\Circled{4} Feature Map}; % / Barrier template
        \node[block, left=.4cm of feature] (optimizer) {\Circled{5} Solver};
        % \node[block, below=of optimizer] (verifier) {Verifier};
        \node[draw, dashed, inner sep=0.25cm, rounded corners,
        % fit=(estimator)(feature)(optimizer)(verifier), label=above:{\textbf{Lucid}}] {};
        fit=(estimator)(feature)(optimizer)(sets)(tuner), label=below:{\textbf{\lucid}}] {};

        % Output block
        % \node[output-block, left=of verifier] (output) {Barrier certificate};
        \node[output-block, below=of optimizer] (output) {Barrier Certificate};
        \node[output-block, below=.5cm of output] (prob) {Safety Probability};
        \node[] at ($(output)!0.5!(prob)$) (plus) {$+$};

>>>>

        % Lucid blocks
        \node[block, below=1cm of input] (estimator) {\Circled{2} Estimator};
        \node[block, right=of estimator] (tuner) {\Circled{3} Tuner};
        \node[block, left=of estimator] (sets) {\Circled{1} Safety Specification};
        \node[block, below=of estimator, text width=2.4cm] (feature) {\Circled{4} Feature Map}; % / Barrier template
        \node[block, left=.4cm of feature] (optimizer) {\Circled{5} Solver};
        % \node[block, below=of optimizer] (verifier) {Verifier};
        \node[draw, dashed, inner sep=0.25cm, rounded corners,
        % fit=(estimator)(feature)(optimizer)(verifier), label=above:{\textbf{Lucid}}] {};
        fit=(estimator)(feature)(optimizer)(sets)(tuner), label={[xshift=3cm]below:{\textbf{\lucid}}}] {};

        % Output block
        % \node[output-block, left=of verifier] (output) {Barrier certificate};
        \node[output-block, below=of optimizer] (output) {Barrier Certificate};
        \node[output-block, right=.5cm of output] (prob) {Safety Probability};
        \node[] at ($(output)!0.5!(prob)$) (plus) {$+$};
